Rapid Cycleway Prioritisation Tool

### 2020-06-11 code/index.Rmd
Add updated landing page text

The [Rapid Cycleway Prioritisation Tool](https://www.cyipt.bike/rapid/) is a research project and web application supported by the [Department for Transport](https://www.gov.uk/government/organisations/department-for-transport) and [Sustrans](https://www.sustrans.org.uk/) to help identify promising locations for road space reallocation in England.
The tool provides an evidence base to support investment in cycling infrastructure, in response to new pressures placed on transport and public health systems by the COVID-19 crisis, at local and regional levels.
The rational for the tool is that, in the context of COVID-19, big decisions around where to prioritise investment on the road network need to be made quickly.
The tool provides a ‘first pass’ assessment of promising locations for new cycleways which have high latent demand for cycling and where road space can be reallocated whilst maintaining two-way traffic flows for motorists.

>>>>

[Sustrans](https://www.sustrans.org.uk/) and the [Department for Transport](https://www.gov.uk/government/organisations/department-for-transport) commissioned ITS Leeds to develop the [Rapid Cycleway Prioritisation Tool](https://www.cyipt.bike/rapid/) to help identifying promising locations for new cycleways in England.
The immediate purpose of this project is to inform bids for Tranche 2 of the [Emergency Active Travel Fund](https://www.gov.uk/government/publications/emergency-active-travel-fund-local-transport-authority-allocations), particularly for authorities which have not yet developed Local Cycling and Walking Investment Plans ([LCWIP](https://www.gov.uk/government/publications/local-cycling-and-walking-infrastructure-plans-technical-guidance-and-tools)). 
Tranche 2 will fund temporary or permanent infrastructure schemes aimed at increasing active travel and helping to shift people away from public transport given capacity constraints imposed by social distancing.

For each combined and local authority, the tool identifies existing cycleways, promising locations for new cycleways on roads with spare space ('top ranked new cycleways') and what a cohesive cycling network might look like in the longer-term.

# How does the tool identify 'top ranked new cycleways'?

The tool identifies priority locations for new cycleways, ranking roads by their ‘cycling potential’ estimated using the Propensity to Cycle Tool ([PCT](https://www.pct.bike/)). This represents the estimated number of cycling trips along this road travelling to or from work or school assuming the Government’s [aim](https://www.gov.uk/government/publications/cycling-and-walking-investment-strategy) to double cycling by 2025 is met.

The 'top ranked new cycleways' represents the roads with the highest cycling potential which also have spare space; that is, are either wide or have two or more road lanes in one direction.
We have restricted the analysis to focusing on roads with spare lanes since these have the capacity to accommodate new cycleways whilst maintaining two-way traffic.
New cycleways on these routes may therefore be faster to deliver than those requiring roads to be closed to traffic or introducing one-way systems.
In addition, these are likely to represent the key arterial routes into town and city centres; hence new cycleways on these roads could result in large increases in cycling volumes.

# How does the tool identify what a 'cohesive network' might look like?

The tool also identifies what a ‘cohesive network’ for cycling might look like if we were to consider a wider range of interventions such as closing roads to motorised traffic and creating one-way systems. This has been estimated by joining up roads that have a high cycling potential on some or all their length. The ‘cohesive network’ might help authorities considering area-wide interventions such as pedestrian and cycle zones or modal filters.

# How can I find the results for my area?

Results for each combined and local authorities can be accessed by clicking on the relevant area from the map below or search the authority’s name from the list below.
This opens a separate webpage presenting a map of the areas identifying ‘existing cycleways’, ‘top ranked new cycleways’ and the ‘cohesive network’.
It also presents a table providing various data about the top ranked new cycleways including the cycling potential, length of continuous cycle network (including the existing network), current speed limit and whether the road is wide and/or has spare road lanes.

# How should I use these results?

Authorities may wish to use results from the tool to help prioritise their bids for Tranche 2 of the Emergency Active Travel Fund, complementing local knowledge and stakeholder engagement.

For example, authorities considering new cycleways may want to review the ‘top ranked new cycleways’ to inform their decision where best to invest. The table below the maps allows users to rank potential new cycleways by various metrics including:

- cycling potential (default metric to rank schemes)
- total length of cycling infrastructure created
- cycling potential x infrastructure length

In addition, the table provides information about road speed limits, width and road lanes which might help users filter out locations which are less desirable for them.
For example:

- Trunk roads with speed limits of 40mph or over may be less suitable for cycleways given their strategic importance for heavy goods vehicles or potentially due to safety concerns for cyclists (particularly for temporary cycleways).
- Authorities may wish to prioritise new cycleways on wide roads rather than reallocating road lanes from motorists to cyclists.

Additionally, authorities considering area-wide measures may wish to consider the data for the ‘cohesive network’. Currently the tool cannot identify specific interventions to promote cycling (e.g. such as pedestrian and cycle zones or modal filters); however, it should give an indication of where such measures might add greatest value.

# What are the key limitations of the tool?

- Currently the tool has been developed primarily to help authorities prioritise the best locations for new cycleways. While it provides some indication of what a cohesive network might look like, it doesn’t seek to prioritise what is the best type of intervention for a given location.
- The ranking of cycleways may not represent a proxy for which schemes are likely to deliver the best value for money. As a consequence this tool doesn’t remove the need to use the Department for Transport’s Active Mode Appraisal Toolkit (AMAT) to assess whether proposed schemes deliver value for money.
- The best locations for new cycleways may not always be on roads with spare lanes. For example, in some circumstances it may be preferable for new cycleways to be along quiet roads running parallel to major arterial routes.
- The results for the ‘top ranked new cycleways’ doesn’t identify routes under 500m with multiple changes of road name and doesn’t identify short routes (under 500m) linking up existing cycleways. To address this limitation, the tool includes results labelled ‘spare lane(s)’ and ‘estimated width >10m’ which plug these gaps. These layers are hidden by default but can be found by clicking the ‘layer’ button in the top left corner of the map.
- The road network data is from 2017 and so will be slightly out-of-date.
- The model estimates the cycling potential of roads based on the estimated number of trips to and from work and school. Currently it is not possible to estimate cycling potential accounting for all journey purposes including shopping and leisure trips, which would significantly increase the estimated cycling potential.
- The PCT uses data from the 2011 Census and 2011 Cycle to School Census to estimate cycling potential. Hence the estimated cycling potential may represent an underestimate if there has been significant construction active in the area since 2011.
- The tool currently only identifies priority on-road routes for new cycleways. There may also be desirable off-road locations for new cycleways including alongside old railway lines or across open areas.
- Currently the tool allows for new cycleways on trunk roads with speeds of 40mph or more. While in some limited cases this may be desirable (e.g. where trunk roads are causing significant air pollution), in many cases it may not be desirable to reallocate road space from trunk roads to new cycleways. To address this, we have identify which top ranked routes have speed limits of 40mph or more in the tables.

# Where can I find further information about the tool?

The technical specification for the Rapid Cycleway Prioritisation Tool can be found here [include hyperlink], providing further detail about the assumptions underlying the tool.

A webinar about the Rapid Cycleway Prioritisation Tool is being held on 17th June 2020 from 10:30-11:45.
Sign-up details can be found [here](https://register.gotowebinar.com/register/3517000554526190862?source=dft) and recording of the webinar will be available on the website after the event.

Data about the tools informing the model can be found here:

- The Propensity to Cycle Tool
- The Cycling Infrastructure Prioritisation Tool
- OpenStreetMap

For expert users, the code underlying the project can be accessed here.
If you would like to support the project with technical input, please provide feedback via the GitHub issue tracker.
Data generated by the project can be downloaded from GitHub (e.g. see GeoJSON files for London [here](https://github.com/cyipt/popupCycleways/tree/gh-pages/release/greater-london)).

The results from this project have been delivered rapidly and should be used alongside local knowledge and stakeholder engagement.

The tool provides an evidence base to support investment in cycling infrastructure, in response to new pressures placed on transport and public health systems by the COVID-19 crisis, at local and regional levels.
The rational for the tool is that, in the context of COVID-19, big decisions around where to prioritise investment on the road network need to be made quickly.
The tool provides a ‘first pass’ assessment of promising locations for new cycleways which have high latent demand for cycling and where road space can be reallocated whilst maintaining two-way traffic flows for motorists.
